---
title: 'Work with and manage your data efficiently in R: tidyr and dplyr'
author: "Elena Dudukina"
date: "26 October 2018"
output: html_document
---

```{r setup, include=T,echo=F}
knitr::opts_chunk$set(echo = TRUE)
```
# Packages to install
```{r install packs, eval=FALSE, include=T}
install.packages(c("tidyr", "dplyr", "stringr", "ggplot2"))
```

# Objectives

1. To learn the difference between "messy" and "tidy" data and see the advantages of using data that are consistently structured
2. To get to know the tools that help tidying up and reshaping the data
3. To perform simple data transformations and create summaries
    
**At the end of this session you will be able:**

1. To change the structure of the data using tidyr tools `gather()`, `separate()`, and `spread()`
2. To perform simple data transformations using dplyr tool `mutate()`
3. To select variables and observations to work with using dplyr tools `filter()`and `select()`
4. To order data by variable using dplyr tool `arrange()`
5. To provide a simple data summary using dplyr tools `group_by` and `summarise()`

# Sources for learning
1. "R for Data Science" book: [Chapter on tidy data](http://r4ds.had.co.nz/tidy-data.html)
2. [tidyr vignette](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)
3. [dplyr vignette](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html)
4. DataCamp course ["Working with Data in the Tidyverse"](https://www.datacamp.com/courses/working-with-data-in-the-tidyverse)
    - Chapter 1 "Explore your data" is free (the following chapters are available after purchasing the DataCamp subscription)
    
# Sources for getting help
1. [Data wrangling cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
2. [StackOverflow for tidyr](https://stackoverflow.com/questions/tagged/tidyr)
3. [StackOverflow for dplyr](https://stackoverflow.com/questions/tagged/dplyr)
4. Quick package information using RStudio: `?tidyr`, `?dplyr`

## "Messy" vs. "tidy" data

**The dataset is tidy if:**

 + Each variable must have its own column
 + Each observation must have its own row
 + Each value must have its own cell

```{r Libraries,}
library(tidyr)
library(dplyr)
library(stringr)
library(ggplot2)
```

Explore the structure of `table1`. Is it tidy?
```{r tab1,}
table1
```
**Pros of tidy data**:

1. Time spent preparing your data at the beginning can save hours of frustration in the long run.
2. "Tidy data" provides a concept and the package tidyr provides the functions for this work.
3. Tibble package makes datasets efficient to work with.
4. dplyr provides fast and intuitive data processing functions; data.table package is speedy with large data.
5. The `%>%` pipe operator can help clarify complex data processing workflows.

## Gather()
Explore the structure of the `table4a`
```{r tab4a, }
table4a
```
* Is `table4a` tidy?
* How would you like to organise data in `table4`?
    - Would it be helpful to have columns `country`, `year`, and `cases` with counts?
    - Gather columns `1999` and `2000` into row values of a new column `year` and cases counts into values of a new column `cases`

```{r tab4a gather,}
table4a_tidy <- table4a %>%
  gather(key = "year", value = "cases", `1999`:`2000`)
table4a_tidy
```
### Exercise 1:
Time: 5 min
```{r tab4b gather,}
# explore structure of `table4b`
table4b

# is it tidy? Use gather() to make it tidy
table4b_tidy <- table4b %>%
  gather (key = "year", value="populaton", `1999`, `2000`)
table4b_tidy
```
## Spread()
Explore the structure of the `table2`
```{r tab2,}
table2
```
* Is `table2` tidy?
    - Depends on what we plan to use it for
    - Later we would like to calculate the rate of the disease in each country per year: rate=cases/100,000*population

* What data structure is preferred?
    - Each country has a row that contains all information needed to proceed with the calculation: year, number of cases, and the population size
    - Spread the column `type` into two columns `cases` and `population`
    
```{r spread,}
table2 %>%
    spread(key = type, value = count)
```

### Case study

The tidyr::who dataset contains tuberculosis (TB) cases broken down by year, country, age, gender, and diagnosis method.

Explore `who` data: are they tidy?
```{r who,}
glimpse(who)
```
`new_sp_m014` to `newrel_f65` are likely values, not variables. We don't know what those values represent yet, so we'll give them the generic name "key". We know the cells represent the count of cases, so we'll use the variable cases. Let's reshape these data:
```{r who1,}
who1 <- who %>% 
  gather(new_sp_m014:newrel_f65, key = "key", value = "cases", na.rm = TRUE)
who1
```

The first three letters of each column denote whether the column contains new or old cases of TB. In this dataset, each column contains new cases.

The next two letters describe the type of TB:

- rel = relapse
- ep = extrapulmonary TB
- sn = smear negative
- sp = smear positive
The sixth letter gives the sex of TB patients. The dataset groups cases by males (m) and females (f).

The remaining numbers gives the age group. The dataset groups cases into seven age groups:

- 014 = 0 - 14 years old
- 1524 = 15 - 24 years old
- 2534 = 25 - 34 years old
- 3544 = 35 - 44 years old
- 4554 = 45 - 54 years old
- 5564 = 55 - 64 years old
- 65 = 65 or older

Fixing inconsistent column names (`new_sp_` but `newrel_`)
```{r who2,}
who2 <- who1 %>% 
  mutate(key = str_replace(string = key, pattern = "newrel", replacement = "new_rel"))
who2
```
Fixing an issue of having several values (e.g., description of the case, type of TB, sex, and age) are coded as one value:
```{r who3,}
who3 <- who2 %>% 
  separate(key, c("new", "type", "sexage"), sep = "_")
who3
```
Dropping redundant columns `new`, `iso2`, `iso3`:
```{r who4,}
who4 <- who3 %>% 
  select(-new, -iso2, -iso3)
str(who4)
```
Separating `sexage` values:
```{r who5,}
who5 <- who4 %>% 
  separate(sexage, c("sex", "age"), sep = 1)
who5
```
### Exercise 2:
Time: 15 min
```{r separate}
# explore the structure of the table3
table3

# separate rate variable into two columns with numeric values.
table3 %>%
  separate(col = rate, into = c("cases", "population"), sep = "/", convert = T) %>%
# calculate new variable "rate" per 100,000 population for each country per year
  mutate(
    rate=as.numeric(cases/population*100000)
  )
```

## Filtering data based on values and ordering data by variables
```{r}
who6 <- who5 %>%
# exclude relapse of TB, include years after 1997 and non-zero case counts
  filter(type != "rel", year > 1997, cases > 0) %>%
# create a new variable called "nordic country"
  mutate(
    nordic = if_else(country == "Denmark" | country == "Sweden" | country == "Norway", "Nordic", "not Nordic")) %>%
# arrange data by year and country
  arrange(year, nordic)
```
### Exercise 3
Time: 15 min
```{r exercise 3,}
# make a dataset who7 that would include only smear positive TB diagnosed from 1994 through 2012
who7 <- who5 %>%
  filter(type == "sp", year >= 1994) %>%
# select all variables except for age
  select(-age) %>%
# arrange by number of cases in descending order
  arrange(desc(cases))

# optional: visualize the case count on y-axis and year on x-axis by sex using fill argument; use ggplot with geom_bar() layer; try dodged position of the bars
who7 %>%
  ggplot(aes(x=year, y=cases, fill=sex)) +
  geom_bar(stat = "identity", position = position_dodge())
```

## Creating a simple data summary
```{r who6,}
summary <- who6 %>%
# group by country
  group_by(country) %>%
# summarise cases count per country
  summarise(
    cum_n_cases = sum(cases),
    pct_fem_cases_cum = sum(cases[sex=="f"])/cum_n_cases*100,
    min_cases = min(cases),
    max_cases =  max(cases)
  ) %>%
  arrange(desc(cum_n_cases)) 
summary
```
## Final exercise:
Time: Until end of the session

1. Choose a dataset to work with
2. Create a new variable using `mutate()`
3. Select variables relevant for the future summary using `select()`
4. Filter data based on several row values using `filter()`
5. Group your data before summarising if necessary using `group_by()`
6. Create a concise summary of the data using `summarise()`
7. Arrange results of your summary using `arrange()`

```{r final exercise example,}
# 1
summary_example <- who6 %>%
# 2
mutate(
  children = if_else( age=="014" & sex=="f", "girls",
                      if_else(age=="014" & sex=="m", "boys", "older"))
) %>%
# 3
select(-sex) %>%
# 4
filter(type == "sn", cases > 30, year > 2000) %>%
# 5
  group_by(year, country) %>%
# 6
  summarise(
    sum_cases = sum(cases),
    n_girl_cases = sum(cases[children=="girls"]),
    pct_girl_cases = sum(cases[children=="girls"])/sum(cases)*100
  ) %>%
# 7
  arrange(desc(pct_girl_cases), sum_cases) 
summary_example
```

