---
title: "Setup and start creating your own package in R"
params:
    datetime: "2018-11-16"
    level: "Intermediate" # or intermediate or advanced
    instructor: "Luke Johnston"
---

```{r setup, include=FALSE}
source("R/utils.R")
library(magrittr)
knitr::opts_chunk$set(comment = "")
```

## Session details

- **Date of session:** `r format_date(params$datetime)`
- **Instructor:** `r params$instructor`
- **Session level**: `r params$level`
    - **Part of the ["Package Development Series"](index.html)**
- **Required packages to install**:

    ```{r, eval=FALSE}
    install.packages(c("devtools", "usethis"))
    ```

### Objectives

1. To see that making packages is not as difficult as it seems.
1. To learn about and use the tools to simplify and automate package development.
1. To provide the resources to expand your knowledge on package development.

**At the end of this session you will be able:**

1. Setup up the necessary folders and files for making a package.
1. Fill out some package description information (in the `DESCRIPTION` and `README.md` files).
1. Include some functions in your package (in the `R/` folder).
1. Install your package.

## Resources for learning and help

- [R Packages](http://r-pkgs.had.co.nz/) online (free) book
- [Creating an R package, a pictorial tutorial](http://www.mjdenny.com/R_Package_Pictorial.html)
- [Package development cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/06/devtools-cheatsheet.pdf)
- [devtools package website](https://devtools.r-lib.org/)
- [usethis package website](https://usethis.r-lib.org/)
- [DataCamp R package course](https://www.datacamp.com/courses/developing-r-packages)
(with a DataCamp subscription)

## List of functions and files covered

**Functions**:

- `usethis::create_package()`
- `usethis::use_r()`
- `usethis::use_readme_md()`
- `usethis::use_mit_license()`
- `devtools::load_all()` (Ctrl-Shift-L)
- `devtools::build()` / `devtools::install()` (Ctrl-Shift-B)
- `usethis::use_usethis()`

**Files**:

- `DESCRIPTION`
- `NAMESPACE`
- `R/`
- `README.md`
- `LICENSE`

## Creating and naming a package

First thing is first: An R package is simply a collection of R functions saved
on your computer that make it easy to access by using the `library()` command.
When you make an R package, it does *not* get uploaded online, it does *not* mean
other people can use it right away. At the beginning, the R package is only available
to you and to your computer. That's it. So why would you want to create a package?
Well, it can save time later with other data analyses since you can access your
functions across files by using `library(<yourpackagename>)`. A package bundles
related code together, simplifying your life! Alright, that's done with, next step!

Creating the initial setup of files and folders necessary for an R package is 
incredibly easy with the usethis package! But! Before we begin, it's a good idea
to think about what the name of the package should be. You can name it whatever
you want, as long as it is not "base", "stats", "graphics", "utils", "tools",
"methods", or any other default base R package. Generally, name it something that
is pretty unique or specific to you. For this session, let's call it by your
initials (or mine "lwj") and add "helpers"... so:

```{r, eval=FALSE}
usethis::create_package("lwj.helpers")
```

```{r, echo=FALSE}
pkg_name <- file.path(tempdir(), "lwj.helpers")
usethis::create_package(pkg_name)
```

A bunch of things are shown on your console. These are just some information
that usethis is telling you that it has done. Otherwise, you now have the beginning
of an R package! Pretty easy eh??

## RStudio project options

Before we get into the package details more, let's make some edits to the project
options. We'll change the "General" options to all "No" and check that the "Build
Tools" has correctly set the project as a package.

## Files and folders

There are many things that make an R package, well, an R package. One of them is
the folder and file structure. All R packages must have certain files and folders
in order to properly install. 

**Required**:

- `R/`: This folder contains the code that is available when the package is
installed. R files should only be in this folder, hence why it is called the
`R/` folder! Covered more below.
- `DESCRIPTION`: Contains all the necessary information about the package, such
as author, version number, name, brief description, and the package dependencies. 
This file is more of the "machine-readable" metadata of the package. Covered more
below.
- `NAMESPACE`: Tells R which functions are available to use from the package. 
At first, this file contains a command to list all functions that are in the `R/` 
folder. We won't cover this file too much during this session.

**Optional, but very helpful**:

- `.Rbuildignore`: This file let's R know what to ignore when you install your
package (for instance, the `.Rproj` file).
- The `.Rproj` file: Let's RStudio know that the directory is an [RStudio
Project](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects)
and that it is a package. Adds some package development functionality to RStudio.

## Package metadata

As mentioned above, there are two main files for providing package "metadata".
Firstly, there is the `DESCRIPTION`, which is required for making a package.
There are several data-fields in the file:

```{r, echo=FALSE}
withr::with_dir(pkg_name, desc::desc())
```

The important data-fields are the `Title`, `Authors`, `Description`, and
`License` (get to this more later). Ok, let's open up the file and make the edits.

> **Exercise** (3-5 min): Open the `DESCRIPTION` file and change the title, author,
and description fields.

The next optional, but important file to have as "human-readable metadata" is the 
`README.md` file. Right now it doesn't exist. But we can create it using:

```{r, eval=FALSE}
usethis::use_readme_md()
```

```{r, echo=FALSE}
withr::with_dir(pkg_name, {
    capture.output(usethis::proj_set(pkg_name))
    usethis::use_readme_md()
})
```

The file looks like this:

```{r, comment="", echo=FALSE}
cat(readr::read_lines(file.path(pkg_name, "README.md")), sep = "\n")
# cat(file = file.path(pkg_name, "README.md"))
```

The `README.md` file is a [markdown](https://rmarkdown.rstudio.com/authoring_basics.html) 
file. We covered [markdown in a previous session](intro-rmarkdown.html).
Generally, this file is a brief documentation containing a high level summary of
what the package does, how to install it, and sometimes a simple example of how
to use it. It should be written and targetted to other (potential) users, or
yourself 9 months in the future (you'll thank yourself). Ok, this file right now 
has some things that need to be changed. Let's edit it.

> **Exercise** (5 min): Write a (fake) goal and description of the package in the
`README.md` file. Remove the text in the "Installation" section for now. Delete
the "Example" section.

Next we want to add the license. The license is really important, even if you 
never put your package online. The license gives information about how other people
can use and modify your code. We won't cover the legal details about licensing,
but for now, I would recommend using an [MIT License](https://en.wikipedia.org/wiki/MIT_License):

```{r, eval=FALSE}
# usethis::use_mit_license("<your name>")
usethis::use_mit_license("Luke Johnston")
```

```{r, echo=FALSE}
withr::with_dir(pkg_name, {
    capture.output(usethis::proj_set(pkg_name))
    usethis::use_mit_license("Luke Johnston")
})
```

The contents of the LICENSE file are:

```{r, comment="", echo=FALSE}
cat(readr::read_lines(file.path(pkg_name, "LICENSE")), sep = "\n")
```

## Adding R code

Now for the main reason you have created a package! To put your R code in it so
you can use the code easily in other scripts! To create an R script in the package,
use:

```{r, eval=FALSE}
# usethis::use_r("<filename>")
usethis::use_r("addition")
```

```{r, echo=FALSE}
# usethis::use_r("<filename>")
withr::with_dir(pkg_name, {
    capture.output(usethis::proj_set(pkg_name))
    usethis::use_r("addition")
})
```

An R script will open in RStudio. Now, let's add a function to it. We'll use a
very simple function in file:

```{r}
addition <- function(x, y) {
    x + y
}
```

Now save. To test that it works, now we run the next command:

```{r, eval=FALSE}
devtools::load_all()
```

Or type "Ctrl-Shift-L". In the console, test if the function works:

```{r, comment="#>"}
addition(2, 2)
```

It should work! We'll add some more functions and play around with the R scripts.

## Building and installing

Now, the final step! Let's build and install the package! When you feel your package
is ready, time to install it for you to use with `library()`! And you can install it
by using...:

```{r, eval=FALSE}
devtools::install()
```

Or by typing "Ctrl-Shift-B". It does some quick checks, and if all goes well, you
now have installed your package!

## Other useful functions

If you end up developing R packages often, I would recommend running this command:

```{r, eval=FALSE}
usethis::use_usethis()
```

This let's you add some code to automatically load when you open up R, to make it
easier to continue your package development!
