---
title: "Setup and start creating your own package in R"
params:
    datetime: "2018-11-16"
    level: "Intermediate" # or intermediate or advanced
    instructor: "Luke Johnston"
---

```{r setup, include=FALSE}
source("R/utils.R")
```

## Session details

- **Date of session:** `r format_date(params$datetime)`
- **Instructor:** `r params$instructor`
- **Session level**: `r params$level`
    - **Part of the ["Package Development Series"](index.html)**
- **Required packages to install**:

    ```{r, eval=FALSE}
    install.packages(c("devtools", "usethis"))
    ```

### Objectives

1. To see that making packages is not as difficult as it seems.
1. To 
1. To learn about and use the tools to develop packages.
1. 

**At the end of this session you will be able:**

1. Setup up the necessary folders and files for making a package.
1. Fill out some package description information.
1. Include some functions in your package.
1. Install your package.

## Resources for learning and help

- [R Packages](http://r-pkgs.had.co.nz/) online (free) book
- [Creating an R package pictorial tutorial](http://www.mjdenny.com/R_Package_Pictorial.html)
- [Package development cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/06/devtools-cheatsheet.pdf)
- [DataCamp R package course](https://www.datacamp.com/courses/developing-r-packages)
(with a DataCamp subscription)

## List of functions and files covered

**Functions**:

- `usethis::create_package()`
- `usethis::use_r()`
- `usethis::use_usethis()`
- `usethis::use_readme_md()`
- `devtools::document()`
- `devtools::build()` / `devtools::install()`

**Files**:

- `DESCRIPTION`
- `NAMESPACE`
- `R/`
- `README.md`

## Creating a package

```{r, eval=FALSE}
pkg_path <- file.path(tempdir(), "testing")
usethis::create_package(pkg_path)
```

```{r}
library(stringr)
fs::dir_ls(pkg_path, all = TRUE, recursive = TRUE)

```


Outline:

- package setup and naming
- RStudio project options
- 
